---
title: "Working with Data"
date: "`r Sys.Date()`"
output:
  rmdformats::html_docco:
    highlight: kate
    code_download: true
---


```{r setup, include=FALSE}
library(knitr)
library(rmdformats)

## Global options
options(max.print="75")
opts_chunk$set(cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```


# Importing Data  

* Use the __Import Dataset__ button in the __Environment__ tab  
* Choose the __From CSV__ option  
* Click on the __Browse__ button  
* Direct the computer to where you saved the BlackfootFish data file, click __open__  
* It will bring up a preview of the data 
* Uncheck the box labeled "Strings as factors"  
* Click on the __Import__ button  

Notice the code that outputs in the console (the bottom left square). This is 
the code that you could have typed in the code chunk below to import the data 
yourself. Copy and paste the code that was output in the code chunk below.  

```{r data, echo = FALSE, message = FALSE, warning = FALSE, purl = TRUE}
library(readr)
BlackfootFish <- read.csv("data/BlackfootFish.csv",
                          stringsAsFactors = FALSE)
# copy and paste the code that was used by R to import the data 
# be careful to only copy the code that is next to the > signs! 
```

\vspace{0.5cm}

```{r fakeData}
# copy and paste the code that was used by R to import the data 
# be careful to only copy the code that is next to the > signs! 
```

# Structure of Data  

The data we will use is organized into data tables. When you imported the 
BlackfootFish data into RStudio was saved as an object. You are able to inspect
the structure of the BlackfootFish object using functions built in to `R` 
(no packages necessary).  

Run the following code. What is output from each of the following commands?

\vspace{0.25cm}

```{r structure}
class(BlackfootFish) ## What is the object class of the data?
dim(BlackfootFish) ## What is the first number represent? What about the second number?
names(BlackfootFish) ## What does this output refer to?
str(BlackfootFish) ## What is the structure of the data?
```

\newpage 

```{r}
summary(BlackfootFish) ## What is the data type of each variable in our dataset?
typeof(BlackfootFish) ## What is the storage mode of a data.frame?
```

\vspace{0.5cm}

When we inspect dataframes, or other objects in `R`, there are some general 
functions that are useful to check the content/structure of the data. 
Here are some:

- size:  
    + `dim(data)`: rows and columns  
    + `nrow(data)`: number of rows  
    + `ncol(data)`: number of columns  


- content:  
    + `head(data)`: first 6 rows  
    + `tail(data)`: last 6 rows  
    + `View(data)`: opens viewer window in separate tab  
    

- names:  
    + `colnames(data)`: column names of dataframe  
    + `rownames(data)`: row names of dataframe  


- summary of content:  
    + `str(data)`: structure of object and information about the columns  
    + `glimpse(data)`: similar information to `str`, but neater output  
    + `summary(data)`: summary statistics for each column  
  
\vspace{0.5cm}

*Note:* Many of these are generic functions that will work on other data types, 
not just on `data.frame`s. 

\newpage

# Dataframes 

What is a dataframe? A dataframe is a type of `R` object and is the *de facto* 
structure of tabular data. You can create dataframes by hand, but most of us do 
not use `R` to input our data by hand. Instead, we import our data using `R` 
commands that read in spreadsheets (`read.csv`, `read_csv`, `read_excel` etc.). 
A dataframe is a set of columns, where each column is a vector. Thus, columns 
have the same data type *within* the column, but potentially different data types *across* columns. 

For example, the columns `trip`, `mark`, and `year` are integers (whole numbers),
`weight` and `length` are numeric (numbers with decimals), and `section` and 
`species` are characters. 


# Extracting Data  

If we were interested in accessing a specific variable in our dataset, we can 
use the `$` command. This command extracts the specified variable (on the right
of the `$` sign) from the dataset. When this is extracted, `R` views the variable
as a vector of entries, which is what the [1:18352] refers to.

\vspace{0.25cm}

```{r extract, tidy = TRUE, purl = TRUE}  
years <- BlackfootFish$year 
## extracts year from the dataset and saves it into a new variable named year
str(years) ## using the new variable (remember case matters!)
## How would you determine how long the vector is?
```

\vspace{0.25cm}

Another method for accessing data in the dataset is using matrix notation 
([row, column]). If you look to your right in the __Environment__ window, you 
notice that RStudio tells you the dimensions of the BlackfootFish data. You can 
(roughly) view the dataset as a matrix of entries, with variable names for each 
of the columns. I could instead use bracket notation to perform the same task as
above, using the following code.  

\vspace{0.25cm}

```{r bracket, purl = TRUE}  
years <- BlackfootFish[, 5] 
## This takes ALL rows of data but only the fifth column
## Same as years <- BlackfootFish[1:18352, 5]
str(years)
```

## Practice:

The following is a preview of the dataframe `df`: 

```{r, echo = FALSE}
df <- data.frame(
  x = c("H", "N", "T", "W", "V"), 
  y = c("May", "Oct", "Mar", "Aug", "Feb"),
  z = c(2010, 2015, 2018, 2017, 2019)
  )
df
```

\vspace{0.25cm}

**Exercise 4:** What would be output if you entered: `df[3, ]`?  

\vspace{0.25cm}

**Exercise 6:** What would you input to get an output of `2015`? Can you 
think of two ways to do it?

\vspace{0.25cm}

**Exercise 7:** How would you pull off only columns `x` and `y`? What about 
pulling off only columns `x` and `z`?

\vspace{0.25cm}

**Exercise 8:** How would you modify the script below, to get an output 
of `[1] 22 24`? 

\vspace{0.25cm}

```{r, eval = FALSE}
s <- c(22, 24, 49, 18, 1, 6)
s[]
```

\vspace{0.25cm}

**Exercise 9:** What would be output if you entered: `s[3, ]`?  

\vspace{0.25cm}

# Changing Data Type  

By default, when building or importing a data frame, the columns that contain 
characters (i.e. text) are coerced (= converted) into factors. Depending on what
you want to do with the data, you may want to keep these columns as character. 
To do so, read.csv() and read.table() have an argument called stringsAsFactors 
which can be set to FALSE. This is what happened when we unchecked the 
"stringsAsFactors" box when importing our data. 

In most cases, it is preferable to set stringsAsFactors = FALSE when importing 
data and to convert as a factor only the columns that require this data type. 
But what is a factor? 

Consider the variables species and section. These variables represent a
broader class of what we call categorical variables. In `R` there are two ways 
to store this information, (1) as a series of `character` strings, or (2) as a 
`factor`. In the early days of coding in `R`, factors were more efficient than 
characters, since you only need to store the level of the factor each 
observation went with. 

While factors are still useful in today's statistical analyses and data 
visualizations, they can be tricky to deal with. When you convert a variable 
to a factor, for many operations you will get different results than for a 
character (McNamara & Horton, 2017). 

In these data section has two levels (Johnsrud and ScottyBrown) and species has
four levels (RBT, WCT, Bull, and Brown). If we want `R` to view these variables
as factors instead of characters, we need to change their data type.   

\vspace{0.25cm}

```{r factors, purl = TRUE}  
unique(BlackfootFish$species) ## tells you the unique values of species
unique(BlackfootFish$section) ## tells you the unique values of section  
BlackfootFish$speciesF <- as.factor(BlackfootFish$species) 
## creates a new variable that is the factor version of species
BlackfootFish$sectionF <- as.factor(BlackfootFish$section) 
## creates a new variable that is the factor version of section
```

\newpage

There is also a function that will allow for you to specify the order of the 
levels of a factor! As we saw before, the `as.factor` function chooses the 
levels alphabetically. Suppose you would like for the species to be in the 
following order: Bull, Brown, RBT, and WCT. 

Using the `factor` function this would look like:

\vspace{0.25cm}

```{r, tidy = FALSE}
BlackfootFish$speciesF <- factor(BlackfootFish$species, 
                                levels = c("Bull", "Brown", "RBT", "WCT"))
```

## Practice:  

**Exercise 11:** Year was saved as an integer data type (1989 - 2006), but we 
may want to consider it to be a categorical variable (a factor). Write the `R` 
code to create a new variable called `yearF` that is a factor of `year` 
(as you did with section and species). 

```{r verify}
```

\vspace{0.25cm}

**Exercise 12:** Now, verify that `yearF` is viewed as a categorical variable, 
with the same levels as `year`. (hint: you have already used three functions 
that would do this for you)  

\vspace{0.25cm}

The issue with factors, lies with if/when you want to change it back to a 
number or character. In the code below I've decided that I don't want year 
to be a factor and want to change it back to numeric. What happens when I use 
the `as.numeric()` function on the yearF variable?  

\vspace{0.25cm}

```{r, echo = FALSE}
BlackfootFish$yearF <- as.factor(BlackfootFish$year)
```

```{r factor_convert, tidy = FALSE}
year_recover <- as.numeric(BlackfootFish$yearF)
ds <- data.frame(original = BlackfootFish$yearF, 
                 recovered = year_recover)
head(ds)
```

# Packages  

As we mentioned previously, `R` has many packages, which people around the world
work on to provide and maintain new software and new capabilities for `R`. 
You will slowly accumulate a number of packages that you use often for a variety
of purposes. In order to use the elements (data, functions) of the packages, you
have to first install the package (only once) and then load the package (every 
time).  

We're going to install a few packages that are often used.  

* Use the __Install__ button in the __Packages__ tab  
* Type in `devtools` and `tidyverse` into the blank line (separated by a comma)  
* Check the __Install dependencies__ box  
* Click on the __Import__ button  

\newpage 

There will be a large amount of output coming out of the console. This output is
`R` trying to download the package(s) you requested by contacting the mirror you
chose for it to use when downloading (I chose Northern Michigan University). 
Once the computer has downloaded the packages, it will tell you that 
"`The downloaded binary packages are in`", followed by the location of the files.  

Now that the files are downloaded, we need to load them in order to use them. 
The following code will load each package, please run it!  

\vspace{0.25cm}

```{r packages}  
library(devtools)
library(tidyverse)
```

\vspace{0.25cm}

Notice that when loading the `tidyverse` package that there is a large amount of
output. This output is telling you all of the other packages that are loaded in 
the `tidyverse` package, as well as the functions in the `tidyverse` package
that overwrite (mask) functions from base `R`.  

This is the process you go through if you ever find packages that you would 
like to use! Often packages that you install will need to be updated. To update
a package you can click on the "Tools" tab, then click on "Check for Package
Updates". This will bring up a window that will list all of the packages that 
have newer versions than what you have. Click on the packages that you wish to 
update, or click on the "Select All" button.  


# Finding Help  

One of the chief reasons for `R`'s religious following is its wonderful 
documentation. If you know a function does what you want (say find the variance),
but are not quite sure how it's spelled, what arguments it takes, or what package
it lives in, don't fret! The `?` and `help()` commands are very powerful. For 
functions, placing the `?` before the name, will tell `R` to search for that 
name in all of the functions, in all of the packages you have installed.  

* If it finds _one_ __identical match__, it will display the help file for that 
function in the Help tab in the bottom-right corner.  
* If it finds _more than one_ __identical match__, it will display the functions,
in their respective packages, that you have to choose from.  
* If it find _no_ __identical match__, it will tell you that "`No documentation 
for ____ in specified packages and libraries:,`" and suggests you use a `??` 
instead.  
    
    + A `??` in front of the function name will search __all of `R`__ for named 
    functions similar to what you typed.  
    + The output will tell you what package the function is in, as well as the 
    function's name (`package::function`).  

\vspace{0.25cm}    

If you would like help on a particular package, say one that you just downloaded,
then you can use the same command(s) to get help.  These commands will load up 
a help page (in RStudio) in the Help pane. Each help page is broken down into 
sections:  

* Description: An extended description of what the function does.   
* Usage: The arguments of the function and their default values.     
* Arguments: An explanation of the object each argument is expecting.  
* Details: Any important details to be aware of.    
* Value: The object the function returns.      
* See Also: Any related functions that may be useful.    
* Examples: Some examples for how to use the function.

# Functions  

In `R` there are both functions that are built in (require no package to be 
loaded), as well as functions that are housed within specific packages. You have
already used a few built in functions to inspect the structure of the 
BlackfootFish data (`str`, `class`, `summary`). As we know, a function 
transforms an input (potentially multiple) into an output. You have to provide
`R` with the inputs (arguments) required for the function to generate an output.
The argument(s) inside a function happen after the `(` symbol. You know an 
object is a function when it is immediately followed by a  `(` and the 
corresponding closing `)` comes after the arguments are complete. The output of 
a function does not have to be numerical and it typically is not a single 
number, it can be a set of things or a dataset.   

Arguments describe the details of what a function is to do. Some functions take 
arguments that are specified by the user, or, if left undeclared, take on 
default values. These arguments are typically given names (as seen in the help
file), but the arguments are assumed to follow the order the function 
expects if they are not named (also stated in the help file). When naming an 
argument, the name of the argument is followed by an `=` sign and then the value
of the argument. Notice that here we are using the `=` to declare what value 
each argument is taking on, we **are not** creating a new variable with that 
value assigned to it.   

Suppose we wanted to create a vector of 10 zeros. To do this, we would use both
the `rep` function:  

\vspace{0.25cm}

```{r rep, purl = TRUE} 
# ?rep
rep(0, times = 10) ## repeating 0 three times
rep(times = 10, 0) ## switching order of arguments
rep(0, 10) ## no named arguments
rep(10, 0) ## not what we wanted!
```

\vspace{0.25cm}

Now let's look over some other functions that are often used: 

\vspace{0.25cm}

```{r functions, purl = TRUE, eval = FALSE} 
mean(BlackfootFish$weight)  ## takes a numerical input, but there are NA's in our data
mean(BlackfootFish$weight, argument here! )  ## add in the argument that removes the NA's
median(BlackfootFish$species) ## gives an error because the input is not the correct data type  
cor(BlackfootFish$length, BlackfootFish$weight) ## takes multiple inputs separated by a comma
## Does cor have an option to remove NA's?
```

\vspace{0.25cm}

As seen in the functions above, some functions have *optional* arguments. If 
they are not specified by the user, then they take on their default value 
(`FALSE` for `na.rm`). These options control the behavior of the functions, 
such as whether it includes/excludes NA values. 


# Cleaning Data  

In many instances, you will deal with data that are not "clean". Based on the 
output we received from the `mean()` function, we know that there are NA's in 
the BlackfootFish data, possibly across a variety of variables. Before we used 
`na.rm` as an option to remove NA's *within* a function, but the `na.omit` 
function takes a dataframe and removes any NA's from that dataset. Based on the
output below, how many rows in the BlackfootFish data have an NA present?  

\vspace{0.5cm}

```{r NA, purl = TRUE, tidy = TRUE}
dim(BlackfootFish) ## gives the dimensions of the dataset in (row, column) format  
dim(na.omit(BlackfootFish)) 
## na.omit takes dataframes, matricies, and vectors and returns the object with incomplete cases removed (NA's removed)
```

\vspace{0.5cm}

__Remark:__ The computer is using an algorithm to return a dataset with no NA 
values anywhere in it. This algorithm goes through every row of the dataset and
(roughly) has the following steps,  

* Inspect the row to see if there is an NA anywhere in that row  
* If there is an NA in that row, the logical (`is.na`) evaluates to TRUE, and 
the row is deleted  
* If there is not any NA's in that row, the logical evaluates to FALSE, and the 
row is retained  
* Once it has stepped through every row, the function outputs the "cleaned" 
dataframe

# Subsetting Data  

If we wish to remove all of the NA's from the dataset, we can simply use the 
`na.omit` command from above. We can save the new "clean" dataset under a new name
(creating a new object) or under the same name as before (replacing the old 
object with the new object).

\vspace{0.25cm}

```{r clean, tidy = TRUE}
BlackfootFish_clean <- na.omit(BlackfootFish) 
## Creates a new dataframe, where the NA's have all been removed 
```

\vspace{0.25cm}

# Data Visualization  

There are many different genres of data graphics, with many different variations
on each genre. Here are some commonly encountered kinds:  

* __scatterplots__: showing relationships between two quantitative variables    
* __distributions__: showing distributions of a single quantitative variable      
* __bar charts__: displaying frequencies or densities of a single categorical 
variable  

## Scatterplots

The main purpose of the scatterplot is to show the relationship between two 
variables across several or many cases. Most often, there is a Cartesian 
coordinate system in which the x-axis represents one variable and the y-axis the
second variable.  

\vspace{0.25cm}

```{r scatter1, tidy = FALSE}
#?plot()
plot(length ~ weight, 
     data = BlackfootFish_clean) 
```

\newpage

```{r scatter2, tidy = FALSE}
plot(length ~ weight, 
     data = BlackfootFish_clean, 
     xlab = "Weight (gm)", ## adding in axis labels
     ylab = "Length (cm)", 
     las = 1 ## changing orientation of axis labels
     ) 
```

## Distribution

A histogram shows how many observations fall into a given range of values of a 
variable.  

\vspace{0.25cm}

```{r hist1, purl = TRUE}
hist(BlackfootFish_clean$length) 
```

\newpage

```{r hist2, tidy = FALSE}
hist(BlackfootFish_clean$length, 
     freq = F) ## converts to a density plot (area adds to 1) 
```

\vspace{0.25cm}

```{r, eval = FALSE}
## Does freq need to be named?
hist(BlackfootFish_clean$length, FALSE)
## Why is there an error about the "number of breaks"?
```

\vspace{0.25cm}

```{r hist3, tidy = FALSE}
hist(BlackfootFish_clean$length, 
     freq = F, 
     xlab = "Length", ## adds x-axis label 
     main = "Fish Lengths in Blackfoot River" ## adds title to plot
     ) 
```

\newpage

```{r hist4, tidy = FALSE}
hist(BlackfootFish_clean$length, 
     freq = F, 
     nclass = 50, ## changes the number of bins 
     xlab = "Length", 
     main = "Fish Lengths in the Blackfoot River", 
     las = 1 ## changes orientation of axis labels
     ) 
```

## Side-by-Side Boxplots  

The familiar boxplot is effective when the objective is to compare the 
distribution of a quantitative variable across different levels of a categorical
variable. 

\vspace{0.25cm}

```{r boxplot, tidy = FALSE, fig.height = 4, fig.width = 5} 
## What other options are available to add to your boxplot?
boxplot(weight ~ species, 
        data = BlackfootFish_clean)
```

\vspace{0.25cm} 

## Bar Charts  

Bar charts are an effective way to compare the frequencies of levels of a 
categorical variable. 

\vspace{0.25cm}

```{r bar, tidy = FALSE, fig.height = 5}
section <- table(BlackfootFish_clean$section)
## tables the number of fish that were caught in each section
barplot(section, 
        xlab = "Section", ## adds axis labels to plot 
        ylab = "Number of Fish", 
        main = "Fish Caught by Section", ## adds title to plot
        las = 1, ## changes orientation of axis labels
        col = "blue", ## specifies what color the bars should be
        ylim = c(0, 12000) ## specifies what range of y-values to plot
        )
```

\vspace{0.25cm}

## Practice  

**Exercise 13:**  Using statistics or graphics, which year in our dataset had the 
most fish caught?  


```{r}
```

\vspace{2cm}

**Exercise 14:** Make a boxplot of the fish weights over the different years in 
the dataset.  

```{r}
```
